Ravi Sood
February 11, 2014
BIOST 578
Homework 2
========================================================

First, load the necessary packages:
```{r}
library(GEOquery)
library(limma)
library(data.table)
```

Next, download the data from GEO and clean up the pData:
```{r}
# Set working directory
setwd("/Users/ravi/Documents/2013-2015 - Graduate School/BIOST 578/")

# Download the data
gds <- getGEO("GSE40812", destdir = "Data/GEO/")
gset <- gds[[1]]
pd <- pData(gset)

# Clean up the pData
keepcols <- c("title", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.2")
pd_clean <- pd[, keepcols]
colnames(pd_clean) <- c("ptid", "cell_type", "status", "treatment")
pd_clean$ptid <- gsub("^.{3,4}_", "", pd_clean$ptid)
pd_clean$ptid <- gsub("\\D", "", pd_clean$ptid)
pd_clean$cell_type <- gsub(".* ", "", pd_clean$cell_type)
pd_clean$status <- gsub(".*: ", "", pd_clean$status)
pd_clean$treatment <- gsub(".*: ", "", pd_clean$treatment)
gset_clean <- gset
pData(gset_clean) <- pd_clean
```

Looking at the macrophage samples only, identify probe sets differentially expressed in response to poly IC treatment:
```{r}
# Create ExpressionSet for macrophage data only
mpd <- pd_clean[pd_clean$cell_type=="Macrophage",]
mset <- gset_clean[, rownames(mpd)]
pData(mset) <- mpd

# Identify differentially expressed probe sets between Poly IC- and mock-treated macrophages
mm_mset <- model.matrix(~treatment, mset)
fit_mset <- lmFit(mset, mm_mset)
ebay_mset <- eBayes(fit_mset)
top_poly <- topTable(ebay_mset, coef="treatmentPoly IC H", number=Inf, sort.by="none")

# Display number of differentially expressed probe sets using FDR of 0.05 and FC > 1.5
sum(top_poly$adj.P.Val<0.05 & abs(top_poly$logFC)>log2(1.5))
```

This returns 1146 poly IC-responsive probe sets. 

Next, look for differential expression of these probe poly IC-responsive sets based on HCV status:
```{r}
# Create expressionSet for poly IC-responsive probe sets only
mset_r <- mset[top_poly$adj.P.Val<0.05 & abs(top_poly$logFC)>log2(1.5),]

# Of the poly IC-responsive probe sets, identify those differentially expressed between HCV+ and HCV- subjects
mm_mset_r <- model.matrix(~treatment+status, mset_r)
fit_mset_r <- lmFit(mset_r, mm_mset_r)
ebay_mset_r <- eBayes(fit_mset_r)
top_poly_r <- topTable(ebay_mset_r, coef="statusPos", number=Inf, sort.by="none")

# As no probe sets are differentially expressed using adjusted p values and FDR <0.05, display the number of differentially expressed probe sets using unadjusted p < 0.05
sum(top_poly_r$P.Value<0.05)
```

This analysis returns 36 poly IC-responsive probe sets differentially expressed between HCV+ and HCV- macrophages. 

Finally, construct a heatmap demonstrating the differential expression of these 36 probe sets among the four groups of samples [mock-treated & HCV(-), mock-treated & HCV(+), poly IC-treated & HCV(-), and poly IC-treated & HCV(+); n = 10/group]:
```{r}
# Create expressionSet for differentially expressed probe sets only
finalset <- mset_r[top_poly_r$P.Value<0.05,]

# Store the column order for heatmap by sorting by 1) treatment, 2) HCV status, and then 3) subject ID
finalpd <- pData(finalset)
finalpd_sorted <- finalpd[order(finalpd$treatment, finalpd$status, finalpd$ptid), ]
hmcolorder <- rownames(finalpd_sorted)

# Store the column names for the heatmap (i.e., ordered subject IDs after sorting first by treatment and then by HCV status)
hmcolnames <- finalpd_sorted$ptid

# Create a matrix of expression values where each row is a probe and each column is a sample
hmdata <- exprs(finalset)

# Sort the columns according to 1) treatment 2) HCV status 3) subject ID
hmdt <- as.data.table(hmdata)
rownames(hmdt) <- rownames(hmdata)
setcolorder(hmdt, hmcolorder)
hmdata <- data.matrix(hmdt)

# Use subject IDs for column names
colnames(hmdata) <- hmcolnames

# Draw the heatmap
heatmap(hmdata, Colv=NA)
```
